\documentclass[11pt,a4paper]{article}

\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{parskip}
\usepackage{natbib}

% ---- notation shortcuts ----
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\kron}{\otimes}
\DeclareMathOperator*{\argmin}{arg\,min}

\title{\textbf{What Happens Inside a Language Model as it Trains?}\\[0.4em]}
\author{Dante Campregher}
\date{\today}

\begin{document}
\maketitle

% =========================================================
\section{What are we trying to understand?}
% =========================================================

When a language model trains, its weights change, but the raw weight matrices are hard to interpret.
A more revealing approach is to look at the \emph{activations}: the vectors that flow through the network when it processes text.
If we collect enough of these vectors, we can ask what their distribution looks like and how that distribution changes over the course of training.

This post describes a set of experiments that do exactly that.
We collect covariance matrices of activations (and gradients) at specific locations inside the network, decompose them into eigenvectors and eigenvalues, and track the resulting eigenvalue spectra from random initialisation through to convergence.

I was inspired by the work of \citet{li2025tracing}, who traced the Entropy of the covariance of final tokens of documents. In this post, we reproduce their work and dig deeper into the training dynamics that cause the macroscopic changes we see. For using K-FAC to estimate curvature in language models, I was inspired by \citet{merullo2025memorization}.

We work with two families of open-weight language models:
\begin{itemize}
  \item \textbf{Pythia}: GPT-style models from 14M to 12B parameters, trained on the Pile.
  \item \textbf{OLMo-2}: 1B and 7B parameter models with a LLaMA-style architecture, trained on OLMo-Mix.
\end{itemize}

Both families release intermediate checkpoints, which is what makes this kind of analysis possible.
Various quantities we use are:
\begin{enumerate}
  \item \textbf{RankMe}: roughly how many directions of the representation space are being occupied by the data.
  \item \textbf{$\alpha$-ReQ}: how quickly the eigenvalue spectrum decays — the power-law exponent.
  \item \textbf{K-FAC log-determinant}: a measure of how curved the loss landscape is near each layer's weights. Specifically: the \emph{log-volume} of the curvature.
  \item \textbf{Mean fraction of total energy}: how much of the uncentered covariance is attributed to the mean of the data.
  \item \textbf{Mean alignment with covariance}: how much the mean lies in the same direction as the uncentered covariance.
\end{enumerate}

The specifics of each will be elaborated on in Section~\ref{sec:metrics}.

% =========================================================
\section{Background: covariance matrices and their spectra}
% =========================================================

\subsection{Covariance matrices from activations}

Consider passing a large batch of text through the model and recording the activation vector $\mathbf{x} \in \R^d$ at some fixed point in the network (say, the input to a particular layer) for every token position.
This gives $N$ vectors $\mathbf{x}_1, \dots, \mathbf{x}_N \in \R^d$, which we stack as rows of a matrix $\mathbf{X} \in \R^{N \times d}$.

The standard way to summarise this cloud of points is through its covariance matrix.
We use two versions:

\paragraph{Uncentered covariance.} The average outer product of the raw activations:
\begin{equation}
  \bm{\Sigma}_A = \frac{1}{N}\mathbf{X}^\top \mathbf{X} \in \R^{d\times d}.
\end{equation}
This captures both the direction of the mean activation and the spread around it.

\paragraph{Centered covariance.} Here we subtract the mean first, isolating the spread:
\begin{equation}
  \hat{\bm{\Sigma}}_A = \bm{\Sigma}_A - \bm{\mu}\bm{\mu}^\top, \qquad \bm{\mu} = \frac{1}{N}\sum_{i=1}^N \mathbf{x}_i.
\end{equation}

The difference between centered and uncentered covariance matters. Namely, If the mean is large, the uncentered covariance will be dominated by one direction, even if the actual variation is distributed across dimensions.

We eigendecompose $\hat{\bm{\Sigma}}_A = \mathbf{V}\bm{\Lambda}\mathbf{V}^\top$, giving eigenvalues $\lambda_1 \ge \lambda_2 \ge \cdots \ge \lambda_d \ge 0$.
The sorted $\lambda$ together make up the \emph{eigenspectrum}, which is the central object of this analysis.

\subsection{Gradient covariance}

Besides forward activations, we also collect \emph{gradient} signals.
For an MLP weight matrix $\mathbf{W}$, the curvature of the loss is usually captured by the Hessian $\mathbf{H} = \nabla_\theta^2 L(\theta)$. Where L is the loss and $\theta$ is the flattened parameters $\mathbf{W}$. The Hessian itself is intractable for any large model, but it can be approximated by the Fisher Information Matrix.
Under the K-FAC approximation \citep{martens2020optimizing}, this factorises as a Kronecker product:
\begin{equation}
  \mathbf{F}_{\mathbf{W}} \approx \bm{\Sigma}_G \kron \bm{\Sigma}_A,
\end{equation}
where $\bm{\Sigma}_A$ is the (uncentered) input activation covariance and $\bm{\Sigma}_G$ is the gradient covariance:
\begin{equation}
  \bm{\Sigma}_G = \frac{1}{N}\mathbf{G}^\top \mathbf{G},
\end{equation}
with $\mathbf{G} \in \R^{N \times d}$ the matrix of per-sample gradients of the loss with respect to the \emph{pre-activation output of the layer}.
$\bm{\Sigma}_A$ tells us which input directions are active; $\bm{\Sigma}_G$ tells us which output directions receive strong gradient signal.
Together, they describe the curvature structure of the loss around each layer's weights.

To collect $\bm{\Sigma}_G$ we use \emph{sampled labels}: we sample from the model's own output distribution and backpropagate through those samples rather than the true labels.
This gives an unbiased estimate of the Fisher information.

% =========================================================
\section{Metrics}\label{sec:metrics}
% =========================================================

\subsection{RankMe: how many dimensions are being used?}

If all the variance in a representation is concentrated in a single direction, the covariance matrix is effectively rank-1.
If variance is spread uniformly across all $d$ dimensions, the matrix has full rank.
RankMe \citep{garrido2023rankme} quantifies where a representation sits between these extremes, using the entropy of the normalised eigenvalue distribution:
\begin{equation}
  p_i = \frac{\lambda_i}{\sum_j \lambda_j}, \qquad
  \text{RankMe} = \exp\!\left(-\sum_i p_i \log p_i\right).
\end{equation}
The result is always between 1 and $d$.
Values near 1 indicate that one eigenvalue dominates; values near $d$ indicate a roughly uniform spectrum.

The eigenvalues $\lambda_i$ are conventionally the eigenvalues of the centered covariance. We follow this convention unless otherwise stated.

The original RankMe paper normalises by singular values $\sigma_i = \sqrt{\lambda_i}$ rather than eigenvalues, which gives slightly different results.
We compute both, but to remain consistent with \citet{li2025tracing}, we focus on the eigenvalue RankMe.

\subsection{Power-law exponent \texorpdfstring{$\alpha$}{alpha}: how fast does the spectrum decay?}

Eigenspectra of neural network layers are rarely flat.
They tend to follow a power law:
\begin{equation}
  \lambda_i \approx C \cdot i^{-\alpha}.
\end{equation}
On a log-log plot this is a straight line with slope $-\alpha$.
Larger $\alpha$ means steeper decay: only a few directions matter.
Smaller $\alpha$ means many dimensions contribute.

We fit $\alpha$ by weighted least-squares following \citet{stringer2019highdimensional}, using weights $w_i = 1/i$ so that the leading (most reliable) eigenvalues carry more influence:
\begin{equation}
  \hat\alpha = \argmin_\alpha \sum_{i} \frac{1}{i} \bigl(\log\lambda_i + \alpha\log i - \log C\bigr)^2.
\end{equation}
% $R^2$ is reported as a quality-of-fit indicator.

\subsection{K-FAC log-determinant}

Under Kronecker-factored Approximate Curvature (K-FAC) Fisher $\mathbf{F}_{\mathbf{W}}$ is approximated as $\bm{\Sigma}_G \kron \bm{\Sigma}_A$, the eigenvalues of the full matrix for a weight $\mathbf{W}$ are the pairwise products $\{\lambda^G_j \lambda^A_i\}$.
A useful scalar summary is the log-determinant, which measures the log-volume of the curvature ellipsoid.

To maintain numerical stability for logs of near-zero eigenvalues. We compute a damped version, adding a small diagonal term proportional to the trace to avoid numerical issues near zero:
\begin{equation}
  \mathcal{L} = d_\text{in} \cdot \log\det(\bm{\Sigma}_G + \varepsilon_G \mathbf{I}) + d_\text{out} \cdot \log\det(\bm{\Sigma}_A + \varepsilon_A \mathbf{I}),
\end{equation}
where $\varepsilon_G = \alpha \cdot \tr(\bm{\Sigma}_G)/d_\text{out}$ and $\varepsilon_A = \alpha \cdot \tr(\bm{\Sigma}_A)/d_\text{in}$ (with $\alpha = 10^{-6}$).

A larger log-determinant indicates stronger curvature around that layer's weights.

% We also examine the joint eigenspectrum directly: the top-20,000 products $\lambda^A_i \lambda^G_j$ sorted in descending order, which shows which input-output direction pairs dominate the curvature.

\subsection{Mean fraction of total energy}

Although RankMe measures the signal around the mean. The weights along with approximations of their curvature depend on the uncentered covariance. To quantify the potential influence of the mean, we track the mean vector $\bm{\mu}$ of the activations alongside the other spectral metrics. When the mean of the activations is large, the distinction between centered and uncentered covariance becomes important.

We track how much of the total signal energy — that is: the uncentered covariance — sits in the mean versus the variation around it:
\begin{equation}
  \rho_\mu = \frac{\|\bm{\mu}\|^2}{\tr(\bm{\Sigma}_A)}.
\end{equation}
When $\rho_\mu$ is close to 1, nearly all the energy is in the mean (the activations cluster tightly around a point far from the origin).
When $\rho_\mu$ is close to 0, the mean is negligible.
% This matters particularly for OLMo-2, where RMSNorm constrains the scale of activations and tends to inflate $\rho_\mu$.

% \subsection{TODO: Hutchinson trace estimator}

% \subsection{Generalised eigendecomposition}

% For layers where we have both $\bm{\Sigma}_G$ (gradient covariance at the output) and $\bm{\Sigma}_B$ (activation covariance at the output), we can ask which output directions have high gradient variance \emph{relative to} their activation variance.
% This is captured by the generalised eigenvalue problem:
% \begin{equation}
%   \bm{\Sigma}_G \mathbf{v} = \lambda \bm{\Sigma}_B \mathbf{v}.
% \end{equation}
% Directions with large $\lambda$ are those where the optimisation signal is large compared to the typical activation, i.e.\ the directions the optimiser is updating most aggressively.

% =========================================================
\section{Experimental setup}
% =========================================================

\subsection{Models and checkpoints}

\begin{table}[ht]
  \centering
  \caption{Model families studied.}
  \label{tab:models}
  \begin{tabular}{llcc}
    \toprule
    Family & Architecture & Sizes we study & Pretraining Set \\
    \midrule
    Pythia & GPT-NeoX & 1B, 6.9B & The Pile deduped \\
    OLMo-2 & LLaMA-style & 1B, 7B & OLMo-Mix \\
    \bottomrule
  \end{tabular}
\end{table}

Both models provide densely logarithmically spaced checkpoints at the beginning, and constantly spaced checkpoints after that.

For both models we sample all densely spaced early checkpoints, and then 20 square-root spaced checkpoints. Resulting in collecting between and 27 and 36 checkpoints across the four models we measure.

\subsection{Where in the network do we collect?}

We hook into three locations:

\paragraph{Final residual stream.}
The vector that feeds into the unembedding head, captured before and after the final layer norm.
This is the model's final embedding representation for the predicted token. When hooking only after the final norm.

\paragraph{MLP projections.}
Inside each transformer block, the MLP contains two or three weight matrices.
Pythia has two (up and down projections); OLMo-2 has three (gate, up, and down).
We collect activation covariances at the input and output of each projection, and gradient covariances at the outputs.

\subsection{Data}

To reproduce the RankMe experiments of \citet{li2025tracing}, we process 5,000 padded sequences between 32 and 512 tokens long from the fineweb dataset. We then sample the final token of each sequence, giving one representation that summarises the sequence.

For our other experiments we sample 6 million tokens (~45.8MB) from packed and shuffled chunks of the respective pretraining mixes of the models. Measuring all but the first 32 tokens for each chunk. Sampling the full token sequence lets us accumulate vastly more sampled gradients economically.

\subsection{Accumulation}

When capturing only the final token, it is economical to store the activation embeddings themselves, but this quickly becomes infeasible when collecting millions of tokens instead of thousands. To do this, we accumulate covariance matrices by summing the rank one projections of the activations. Storing a matrix of size $d \times d$ where d is the embedding dimension at that hook. This can be cheap in the residual stream. But inside MLPs, the embedding dimension is often 4 times the residual's dimension, presenting a tricky storage-compute dilemma if experiments need to be recomputed.

Covariance matrices are accumulated in \texttt{float64} for numerical stability after millions of accumulations; activations are cast from lower precision to \texttt{float32} before forming outer products.

% % =========================================================
% \section{Experiments}
% % =========================================================

% \subsection{Final residual stream}

% The most basic question is how the final output representation changes shape over training.
% We track RankMe, trace (total signal energy), and mean norm across all checkpoints, comparing:
% \begin{itemize}
%   \item Centered vs.\ uncentered covariance (does the mean carry most of the energy?).
%   \item Training set vs.\ FineWeb (are these properties distribution-specific?).
%   \item Before vs.\ after the final layer norm (how much does normalisation change the geometry?).
% \end{itemize}

% The main models are Pythia-1B, Pythia-6.9B, OLMo-2-1B, and OLMo-2-7B.

% \subsection{K-FAC}

% We focus on late-layer MLP projections (layer 15 for 1B models, layer 23 for 7B), where the gradient signal tends to be strongest and most structured.
% For each we plot trace, log-determinant, RankMe, and peak eigenvalue of $\bm{\Sigma}_A$, $\bm{\Sigma}_G$, and the joint K-FAC spectrum.

% \subsection{Behaviour of MLP layers}

% We sweep across all blocks and study the spectral properties at each stage of the MLP:
% \begin{itemize}
%   \item Up/gate projection inputs: the residual stream entering the block.
%   \item Up/gate projection outputs: the post-nonlinearity features (GELU or SiLU).
%   \item Down-projection inputs: the combined gate $\times$ up features (OLMo-2) or the GELU output (Pythia).
%   \item Down-projection outputs: the MLP's contribution back to the residual stream.
% \end{itemize}

% For each of these we report log-det, peak eigenvalue, RankMe, mean fraction $\rho_\mu$, and the full eigenvalue histogram.

% \subsection{Gradient covariances}

% Using sampled-label backpropagation we collect $\bm{\Sigma}_G$ at all MLP projections.
% This allows us to study the full K-FAC structure by combining activation covariances from one projection with gradient covariances from another, forming the joint curvature spectrum.
% We also produce layer-wise heatmaps of $\alpha$ and RankMe to see whether layers specialise differently over training.

% \subsection{Residual stream gradients}

% We also collect gradient covariances at the final residual stream.
% This shows how the gradient signal reaching the output changes over training, and how it interacts with the final layer norm.

% =========================================================
\bibliographystyle{plainnat}
\bibliography{citations}
% =========================================================

\end{document}
